Slitasje og kjørespor

### 2023-01-23 DRAFT_slitasjeOgKjørespor.Rmd
feat: Subset wetlands and validate

```{r setup, include=FALSE}
library(knitr)
library(sf)
library(tmap)
library(ggplot2)
library(units)
library(eaTools)
library(tidyverse)
library(tidyr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

@@

I now have 20 unique areas that I will, given there is data, extrapolate
indicator values over:

```{r}
HIA_regions <- HIA_regions %>%
  mutate(region_HIF = paste(region, infrastructureIndex))
length(unique(HIA_regions$region_HIF))
```

Here is a view of the data zooming in on Trondheim

@@

```{r}
tm_shape(HIA_trd)+
  tm_polygons(col = "infrastructureIndex",
    title="Infrastructure index\n(modified 4-step scale)",
    palette = "viridis",
    style="cat")+
  tm_layout(legend.outside = T)+
  tm_shape(hw_utm)+
  tm_lines(col="blue")
```

Let's calculate the areas of these polygons and compare the HIF in the
five regions. 
```{r}
HIA_regions$area <- sf::st_area(HIA_regions)
```

@@

<!-- Check with updated ea_spread -->
```{r}
mySummary <- ea_spread(indicator_data = slitasje_data,
                         indicator = i_2020,
                         regions = HIA_regions,
                         groups = region_HIF,
                         summarise = TRUE)

```

```{r, eval=F}
slitasje_data_extrapolated <- ea_spread(indicator_data = slitasje_data,
                         indicator = i_2020,
                         regions = HIA_regions,
                         groups = region_HIF,
                         threshold = 20)
#saveRDS(slitasje_data_extrapolated, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/slitasje_data_extrapolated.rds")
```

@@

```{r}
myBB <- st_bbox(c(xmin=260520.12, xmax = 278587.56,
                ymin = 7032142.5, ymax = 7045245.27),
                crs = st_crs(slitasje_data_extrapolated))
```

@@

```{r}
slitasje_data_extrapolated_trd <- sf::st_crop(slitasje_data_extrapolated, myBB)
```

```{r}
tm_shape(slitasje_data_extrapolated_trd)+
  tm_polygons(col = "meanIndicatorValue",
    title="Slitasje indicator",
    palette = "viridis",
    style="cat")+
  tm_layout(legend.outside = T)+
  tm_shape(hw_utm)+
  tm_lines(col="blue")
```

Let's split the ID column to get the HIF

```{r}
slitasje_data_extrapolated <- slitasje_data_extrapolated %>%
  tidyr::separate(ID, into = c("region", "HIF"),
                  sep = " ")
slitasje_data_extrapolated$HIF <- as.numeric(slitasje_data_extrapolated$HIF)
```

Now let's check the correlation between the HIF and the indicator
values.
<!-- This is not correct. This count the polygins that have very different sizes -->

```{r slitasje-box, fig.cap="The relationship between human impact factor (a categorical version of the Norwegian Infrastructure Index) and the indicator avleus for slitasje."}
ggplot(slitasje_data_extrapolated, aes(x = factor(HIF), y = meanIndicatorValue))+
  geom_boxplot(fill = "grey",
               size=1.5)+
  theme_bw(base_size = 12)+
  labs(x = "Human Impact Factor", y = "Slitasje indicator value")
```

```{r}
cor(slitasje_data_extrapolated$HIF, slitasje_data_extrapolated$meanIndicatorValue)
```

@@

```{r, eval=F}
slitasje_data_extrapolated2 <- ea_spread(indicator_data = slitasje_data_extrapolated,
                         indicator = meanIndicatorValue,
                         regions = regions,
                         groups = region,
                         threshold = 20)
saveRDS(slitasje_data_extrapolated2, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/slitasje_data_extrapolated2.rds")
```

```{r, echo=F}
# silently read cached file
slitasje_data_extrapolated2 <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/slitasje_data_extrapolated2.rds")
```

```{r}
plot(factor(slitasje_data_extrapolated2$ID), slitasje_data_extrapolated2$meanIndicatorValue,
     ylab="indicator value", xlab="", ylim=c(0,1))
```

>>>>

```{r setup, include=FALSE}
library(knitr)
library(sf)
library(tmap)
library(ggplot2)
library(DT)
library(units)
library(eaTools)
library(tidyverse)
library(tidyr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

I want to check that HIF is in fact a good predictor for _slitasje_.

@@

```{r}
HIA_regions <- HIA_regions %>%
  mutate(region_HIF = paste(region, infrastructureIndex))
```

#### Validate
I now have 20 unique areas that I will, given there is data, extrapolate
indicator values over:
<!-- NEEDS CACHING -->

```{r, eval=F}
corrCheck <- st_intersection(slitasje_data, HIA_regions)

corrCheck <- as.data.frame(corrCheck) %>%
  group_by(hovedøkosystem, region) %>%
  summarise(cor = cor(i_2020, infrastructureIndex))
```

```{r, fog.cap="Pearson correaltions between the indicator values at the polygons level and the HIF."}

ggplot(corrCheck, aes(x = hovedøkosystem, 
                      y = cor))+
  geom_bar(stat = "identity")+
  facet_wrap(.~region)+
  coord_flip()
```

```{r}
ggarrange(
ggplot(corrCheck, aes(x = hovedøkosystem, 
                      y = cor))+
  geom_bar(stat = "identity")+
  coord_flip(),
ggplot(corrCheck, aes(x = region, 
                      y = cor))+
  geom_bar(stat = "identity")+
  coord_flip(),
ncol=1
)
```

From the figures above it appears that the relationship is quite week in general. But the sheer number of data points suggest that all group differences will be very reliable.
_Naturlig åpne områder_ has the lowest correlation between the indicator and the homogeneous area classes, and of the five regions, _Nord-Norge_ has the lowest correlation.

#### Looking at the HIA
Here is a view of the data zooming in on Trondheim

@@

```{r, fig.cap="A closer look at the HIA designation over Trondheim"}
tm_shape(HIA_trd)+
  tm_polygons(col = "infrastructureIndex",
    title="Infrastructure index\n(modified 4-step scale)",
    palette = "viridis",
    style="cat")+
  tm_layout(legend.outside = T)+
  tm_shape(hw_utm)+
  tm_lines(col="blue")
```

Let's calculate the areas of these polygons and compare the HIF in the
five regions.

```{r}
HIA_regions$area <- sf::st_area(HIA_regions)
```

@@

I will focus on wetland from here on out. Only for open wetlands do we have good exosystem delineation maps to base a spatial averaging on.

```{r}
wetlands <- slitasje_data[slitasje_data$hovedøkosystem == "Våtmark",]
```

I want to add a threshold so that we don't end up over extrapolating
based on too few data points. I think we need a minimum of 100 data point to base an estimate on.

<!-- NEEDS CACHING -->
```{r}
mySummary <- ea_spread(indicator_data = wetlands,
                         indicator = i_2020,
                         regions = HIA_regions,
                         groups = region_HIF,
                         summarise = TRUE,
                       threshold = 100)
```

<!-- Do we need another aggregation step here, to get regional estimates= -->

Let's split the regions column

```{r}
mySummary <- mySummary %>%
  tidyr::separate(region_HIF, into = c("region", "HIF"),
                  sep = " ")
mySummary$HIF <- as.numeric(mySummary$HIF)
```

```{r, fig.cap="Indicator values (area weighted averages) for each region and homogeneous area class, as well as the number of data points per groups."}
ggarrange(
ggplot(mySummary, aes(x = HIF, y = area_weighted_mean_indicator_value,
                      fill = area_weighted_mean_indicator_value))+
  geom_bar(stat = "identity")+
  theme_bw(base_size = 12)+
  facet_wrap(.~region, ncol=1)+
  guides(fill = "none"),
ggplot(mySummary, aes(x = HIF, y = data_points,
                      fill = region))+
  geom_bar(stat = "identity")+
  theme_bw(base_size = 12)+
  facet_wrap(.~region, ncol=1)+
  guides(fill = "none")
)
```

The figure above shows that there are much more data on the two middle
classes (class 1 and 2). We do not have enough data to extrapolate to the 
highest impacted areas in any of the regions. These are not the areas prioritised for nature type mapping.

We also see that, even though there are small differences, high HIF is assiciated with lower indicator values.

```{r}
DT::datatable(mySummary)
```

```{r, eval=F}
wetlands_slitasje_extr <- ea_spread(indicator_data = wetlands,
                         indicator = i_2020,
                         regions = HIA_regions,
                         groups = region_HIF,
                         threshold = 100)
#saveRDS(wetlands_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr.rds")
```
```{r, include=F}
wetlands_slitasje_extr <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr.rds")
```

@@

```{r}
myBB <- st_bbox(c(xmin=260520.12, xmax = 278587.56,
                ymin = 7032142.5, ymax = 7045245.27),
                crs = st_crs(wetlands_slitasje_extr))
```

@@

```{r}
wetlands_slitasje_extr_trd <- sf::st_crop(wetlands_slitasje_extr, myBB)
```

```{r, fig.cap="Sliatsje (wetland) indicator extrapolated over Trondheim"}
tm_shape(wetlands_slitasje_extr_trd)+
  tm_polygons(col = "area_weighted_mean_indicator_value",
    title="Slitasje indicator",
    palette = "viridis",
    style="cont",
    breaks = c(0,1))+
  tm_layout(legend.outside = T)+
  tm_shape(hw_utm)+
  tm_lines(col="blue")
```

Let's split the ID column to get the HIF and plot agains indicator values

```{r}
wetlands_slitasje_extr <- wetlands_slitasje_extr %>%
  tidyr::separate(ID, into = c("region", "HIF"),
                  sep = " ")
wetlands_slitasje_extr$HIF <- as.numeric(wetlands_slitasje_extr$HIF)
```

```{r}
summary(wetlands_slitasje_extr)
ggplot(wetlands_slitasje_extr, aes(x = factor(HIF), y = area_weighted_mean_indicator_value))+
         geom_boxplot(fill = "grey")+
  theme_bw(base_size = 12)
```

@@

```{r, eval=F}
wetlands_slitasje_extr2 <- ea_spread_dev(indicator_data = wetlands_slitasje_extr,
                         indicator = area_weighted_mean_indicator_value,
                         regions = regions,
                         groups = region,
                         summarise = F)
saveRDS(wetlands_slitasje_extr2, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr2.rds")
```

```{r, include=F}
# silently read cached file
wetlands_slitasje_extr2 <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr2.rds")
```

```{r}
plot(factor(wetlands_slitasje_extr2$ID), wetlands_slitasje_extr2$area_weighted_mean_indicator_value,
     ylab="indicator value", xlab="", ylim=c(0,1))
```
